The state learner -- a super learner for right-censored data

### 2024-05-17 tex/jrss-version/state-learner-jrss-style.tex
final read through

In both parts we use the same data-generating mechanism. We generate
data according to a distribution motivated from a real dataset in
which censoring depends on the baseline covariates. We simulate data
based on the prostate cancer study of
\cite{kattan2000pretreatment}. The outcome of interest is the time to
tumor recurrence, and five baseline covariates are used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
(HT, yes/no) and clinical stage (CS, six values). The study was
designed such that a patient's radiation dose depended on when the
patient entered the study \citep{gerds2013estimating}. This in turn
implies that the time of censoring depends on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a
sensitivity analysis was conducted based on simulated data. Here we use the
same simulation setup, where event and censoring times are generated
according to parametric Cox-Weibull models estimated from the original
data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original
data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
simulation setting as ``dependent censoring''. We also considered a
simulation setting where data were generated in the same way, except
that censoring was generated completely independently. We refer to
this simulation setting as ``independent censoring''.

@@

We use the learners of the three cumulative hazard functions selected
by the state learner and the estimator defined in
Section~\ref{sec:cause-spec-aver},
equation~(\ref{eq:one-step-comp-ate}), to estimate the average
treatment effect of hormone therapy on risk of tumor recurrence and
death. The propensity score is estimated with a lasso model that
includes all levels of interaction.
% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
%         der mangler en reference for lasso model
The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
month intervals after baseline with point wise 95\% confidence
intervals. We see that hormone therapy decreases the risk of tumor
recurrence and increases the risk of death without tumor recurrence,
but that none of the estimated effects are statistically significant.

\begin{figure}
  \centering%
  \includeFigCond[width=.8\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

>>>>

In both parts we use the same data-generating mechanism. We generate
data according to a distribution motivated from a real data set in
which censoring depends on the baseline covariates. We simulate data
based on the prostate cancer study of
\cite{kattan2000pretreatment}. The outcome of interest is the time to
tumor recurrence, and five baseline covariates are used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
(HT, yes/no) and clinical stage (CS, six values). The study was
designed such that a patient's radiation dose depended on when the
patient entered the study \citep{gerds2013estimating}. This in turn
implies that the time of censoring depends on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a
sensitivity analysis was conducted based on simulated data. Here we use the
same simulation setup, where event and censoring times are generated
according to parametric Cox-Weibull models estimated from the original
data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original
data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
simulation setting as ``dependent censoring''. We also considered a
simulation setting where data were generated in the same way, except
that censoring was generated completely independently. We refer to
this simulation setting as ``independent censoring''.

@@

We use the learners of the three cumulative hazard functions selected
by the state learner and the estimator defined in
Section~\ref{sec:cause-spec-aver},
equation~(\ref{eq:one-step-comp-ate}), to estimate the average
treatment effect of hormone therapy on risk of tumor recurrence and
death. The propensity score is estimated with a lasso model that
includes all levels of interaction.
% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
%         der mangler en reference for lasso model
The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
month intervals after baseline with pointwise 95\% confidence
intervals. We see that hormone therapy decreases the risk of tumor
recurrence and increases the risk of death without tumor recurrence,
but that none of the estimated effects are statistically significant.

\begin{figure}
  \centering%
  \includeFigCond[width=.8\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}
